---
title: "TSS profiles: plotting & clustering"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
options(stringsAsFactors = FALSE)
knitr::opts_chunk$set(echo = TRUE)
library(GenomicRanges)
library(rtracklayer)
library(tidyverse)
source("../../util/plotting_functions.R")
source("../../util/intersect_functions.R")
source("../../util/_setup.R")
```

# TSS profile plots

## Import the data we need

We'll be using the consensus peaks we generated previously (MACS peak calls filtered to those that are overlapping across replicates) and the Gencode gene annotations. 

```{r import}
gencode_gr <- rtracklayer::import("/scratch/Shares/rinn/genomes/Mus_musculus/Gencode/M25/gencode.vM25.annotation.gtf")
fl <- list.files("/scratch/Shares/rinn/JR/FIRRE_KO_TG_K27me3_CHIPSEQ/analysis/02_meta_plots/results/broad_peaks", full.names=TRUE)

consensus_peaks <- import_peaks("/scratch/Shares/rinn/JR/FIRRE_KO_TG_K27me3_CHIPSEQ/results/broad_peaks")


names(consensus_peaks) <- sapply(consensus_peaks, function(x){
  unlist(strsplit(x$name, "/"))[[4]]
})

num_peaks_df <- data.frame("dbp" = names(consensus_peaks),
                           "num_peaks" = sapply(consensus_peaks, length))

write.csv(num_peaks_df, "results/num_peaks_df.csv")

?coverage

```

## Step 1: Create promoter windows

Here we're going to create a GRanges object that contains 6kb promoter windows for each gene in the Gencode annotation. First we'll need to filter the Gencode GRanges object to just the genes and then we can use the promoter function from GRanges that will allow us to specify how big of a window we want upstream and downstream of the TSS (you can have asymmetrical windows).

```{r}
# Recall: The gencode annotation contains an entry for each exon, transcript, etc.
# Use table(gencode_gr$type) to see how many of each entry there are. 
# We don't want to create a "promoter" window around each exon for example
# which is why we need to filter to just the genes.
genes <- gencode_gr[gencode_gr$type == "gene"]

# This function is a convenience function built into GenomicRanges
all_promoters_gr <- promoters(genes, upstream = 3e3, downstream = 3e3)

summary(all_promoters_gr)
table(width(all_promoters_gr))
```

## Step 2: Transform peaks into a coverage object

In order to calculate what the peak coverage across each promoter is we'll convert the peaks GRanges object which currently holds a range for each peak into a run-length encoded list where 0 represents the genomic coordinates where there is no peak present and 1 represents the locations where a peak is present. The reason for run length encoding is that storing this vector without this compression means that we would be storing a numeric value for each of the 3.2 billion base pairs. This would mean allocating a vector in memory that's ~180 GB -- instead with run-length encoding we're in the ~100 KB range. 


# Let's put all this into a funciton so we can call it later conveniently.

```{r}
profile_tss <- function(peaks, 
                        promoters_gr,
                        upstream = 3e3,
                        downstream = 3e3) {
  
  peak_coverage <- coverage(peaks)
  
  coverage_length <- elementNROWS(peak_coverage)
  coverage_gr <- GRanges(seqnames = names(coverage_length),
                         IRanges(start = rep(1, length(coverage_length)), 
                                 end = coverage_length))
  
  promoters_gr <- subsetByOverlaps(promoters_gr, 
                                   coverage_gr, 
                                   type="within", 
                                   ignore.strand=TRUE)
  chromosomes <- intersect(names(peak_coverage), 
                           unique(as.character(seqnames(promoters_gr))))
  peak_coverage <- peak_coverage[chromosomes]
  
  promoters_ir <- as(promoters_gr, "IntegerRangesList")[chromosomes]
  
  promoter_peak_view <- Views(peak_coverage, promoters_ir)
  
  promoter_peak_view <- lapply(promoter_peak_view, function(x) t(viewApply(x, as.vector)))
  promoter_peak_matrix <- do.call("rbind", promoter_peak_view)
  
  minus_idx <- which(as.character(strand(promoters_gr)) == "-")
  promoter_peak_matrix[minus_idx,] <- promoter_peak_matrix[minus_idx,
                                                           ncol(promoter_peak_matrix):1]
  
  promoter_peak_matrix <- promoter_peak_matrix[rowSums(promoter_peak_matrix) > 1,]
  
  peak_sums <- colSums(promoter_peak_matrix)
  peak_dens <- peak_sums/sum(peak_sums)
  
  metaplot_df <- data.frame(x = -upstream:(downstream-1),
                            dens = peak_dens)
  
  return(metaplot_df)
}
```

# Run the function

```{r}
# Here's how we'd run the function for one sample.
KO_CTL_48_IP_R1_peak_4_metaplot_profile <- profile_tss(consensus_peaks[["KO_CTL_48_IP_R1_peak_4"]], all_promoters_gr)

ggplot(KO_CTL_48_IP_R1_peak_4_metaplot_profile, 
       aes(x = x, y = dens)) +
  geom_vline(xintercept = 0, lty = 2) + 
  geom_line(size = 1.5) + 
  ggtitle("Promoter Metaplot") + 
  scale_x_continuous(breaks = c(-3000, 0, 3000),
                     labels = c("-3kb", "TSS", "+3kb"),
                     name = "") + 
  ylab("Peak frequency") + 
  scale_color_manual(values = c("#424242","#a8404c"))
```

Next sample ....
```{r}
WTiTg_48_IP_R1_peak_4_metaplot_profile <- profile_tss(consensus_peaks[["WTiTg_48_IP_R1_peak_4"]], all_promoters_gr)

ggplot(WTiTg_48_IP_R1_peak_4_metaplot_profile, 
       aes(x = x, y = dens)) +
  geom_vline(xintercept = 0, lty = 2) + 
  geom_line(size = 1.5) + 
  ggtitle("Promoter Metaplot") + 
  scale_x_continuous(breaks = c(-3000, 0, 3000),
                     labels = c("-3kb", "TSS", "+3kb"),
                     name = "") + 
  ylab("Peak frequency") + 
  scale_color_manual(values = c("#424242","#a8404c"))

```

```{r}

metaplot_profiles <- lapply(consensus_peaks, profile_tss, promoters_gr = all_promoters_gr)
for(i in 1:length(metaplot_profiles)) {
  metaplot_profiles[[i]]$sample_name <- names(consensus_peaks)[[i]]
}
metaplot_profiles_df <- bind_rows(metaplot_profiles)

ggplot(metaplot_profiles_df %>% filter(grepl("WT", sample_name)), 
       aes(x = x, y = dens, color = sample_name)) +
  geom_vline(xintercept = 0, lty = 2) + 
  geom_line(size = 1.5) + 
  ggtitle("Promoter Metaplot") + 
  scale_x_continuous(breaks = c(-3000, 0, 3000),
                     labels = c("-3kb", "TSS", "+3kb"),
                     name = "") + 
  ylab("Peak frequency") 



sapply(consensus_peaks, length)
# Plotting


```

```{r}
ggplot(metaplot_profiles_df %>% filter(sample_name %in% c("KO_CTL_48_IP_R1_peak_4", "KO_CTL_IP_R1_peak_4")), 
       aes(x = x, y = dens, color = sample_name)) +
  geom_vline(xintercept = 0, lty = 2) + 
  geom_line(size = 1.5) + 
  ggtitle("Promoter Metaplot") + 
  scale_x_continuous(breaks = c(-3000, 0, 3000),
                     labels = c("-3kb", "TSS", "+3kb"),
                     name = "") + 
  ylab("Peak frequency") 
```

```{r}
ggplot(metaplot_profiles_df %>% filter(sample_name %in% c("KOiTg_48_IP_R1_peak_4", "KO_CTL_48_IP_R1_peak_4")), 
       aes(x = x, y = dens, color = sample_name)) +
  geom_vline(xintercept = 0, lty = 2) + 
  geom_line(size = 1.5) + 
  ggtitle("Promoter Metaplot") + 
  scale_x_continuous(breaks = c(-3000, 0, 3000),
                     labels = c("-3kb", "TSS", "+3kb"),
                     name = "") + 
  ylab("Peak frequency") 
ggsave("figures/metaplot.pdf")
```

```{r}
ggplot(metaplot_profiles_df %>% filter(sample_name %in% c("WT_CTL_48_IP_R1_peak_4", "WTiTg_48_IP_R1_peak_4")), 
       aes(x = x, y = dens, color = sample_name)) +
  geom_vline(xintercept = 0, lty = 2) + 
  geom_line(size = 1.5) + 
  ggtitle("Promoter Metaplot") + 
  scale_x_continuous(breaks = c(-3000, 0, 3000),
                     labels = c("-3kb", "TSS", "+3kb"),
                     name = "") + 
  ylab("Peak frequency") 
```

```{r}
ggplot(metaplot_profiles_df %>% filter(sample_name %in% c("WT_CTL_IP_R1_peak_4", "WT_CTL_48_IP_R1_peak_4")), 
       aes(x = x, y = dens, color = sample_name)) +
  geom_vline(xintercept = 0, lty = 2) + 
  geom_line(size = 1.5) + 
  ggtitle("Promoter Metaplot") + 
  scale_x_continuous(breaks = c(-3000, 0, 3000),
                     labels = c("-3kb", "TSS", "+3kb"),
                     name = "") + 
  ylab("Peak frequency") 
```

```{r}
ggplot(metaplot_profiles_df %>% filter(sample_name %in% c("WT_CTL_IP_R1_peak_4", "KO_CTL_IP_R1_peak_4")), 
       aes(x = x, y = dens, color = sample_name)) +
  geom_vline(xintercept = 0, lty = 2) + 
  geom_line(size = 1.5) + 
  ggtitle("Promoter Metaplot") + 
  scale_x_continuous(breaks = c(-3000, 0, 3000),
                     labels = c("-3kb", "TSS", "+3kb"),
                     name = "") + 
  ylab("Peak frequency") 


```

Peaks on promoters
```{r}

promoter_peaks <- lapply(consensus_peaks, function(x) { 
  subsetByOverlaps(x, all_promoters_gr)
})

summary(promoter_peaks)
```





